---
title: "week06-inclass"
author: "Patrick Bitterman"
date: "2025-02-16"
output: html_document
---


## Load packages
```{r}
library(tidyverse)
library(sf)
library(tmap)
library(gstat)
library(SpatialKDE)
library(stars)
```
Note, examples are modified from the R Spatial Data Science book chapter on interpolation. If you want to know/read more, you can do so here: https://r-spatial.org/book/12-Interpolation.html


## read the files in
```{r}
ca.path <- "./data/ca_points/california.gpkg"
ks.path <- "./data/ca_points/ca_k_pbe.gpkg"

ca <- sf::read_sf(ca.path)
ks <- sf::read_sf(ks.path)

```


## ESDA

What are the data?
```{r}
glimpse(ks)


ks %>% 
  ggplot(., aes(x = pbe1415)) + 
  geom_histogram(bins = 40) +
  theme_minimal()

```

What other ESDA could you perform?


## Plot

Let's do some simple mapping as we would normally

```{r}
tm_shape(ca) + tm_polygons() + tm_shape(ks) + tm_dots(fill='navyblue', size = 0.1)

## Note the above controls on point fill and size. Experiment with different values

```


A more interactive way of using TMAP
```{r}
tmap_mode("view")

tm_shape(ca) + tm_polygons() + tm_shape(ks) + tm_dots(fill='navyblue', size = .4)
```

set it back to plot mode

```{r}
tmap_mode("plot")

tm_shape(ca) + tm_polygons() + tm_shape(ks) + tm_dots(fill='navyblue', size = .2)
```

### Dots vs. bubbles
In the tmap R package, both `tm_dots()` and `tm_bubbles()` are used to visualize point data, but they differ in how they represent attributes:

`tm_dots()`: Plots points as uniform-sized dots, useful for simple point visualizations where all points are given equal emphasis.

`tm_bubbles()`: Plots points as circles (bubbles) with variable sizes based on an attribute, allowing visualization of differences in magnitude.

_Key Difference_: `tm_dots()` is for uniform point representation, while `tm_bubbles()` uses point size to convey quantitative differences in an attribute.



# Interpolation

Based on your readings, what's the purpose of interpolation and how does it work???

JARGON AHOY!!!

KDE vs. IDW vs. KRIGING... ANY IDEAS?


### KDE

Kernel Density Estimation (KDE) is a non-parametric spatial analysis technique used to estimate the probability density function of point data over a continuous surface, effectively identifying areas of high and low concentration.


There are many different packages that can calculate a KDE. We're going to use `SpatialKDE` as an example

```{r}
# Define some initial parameters - what are the units???
# Meters because that is the projection Dr. Bitterman used

cell_size <- 5000
band_width <- 10000
## These are probably larger than we'd typically want
## But let's keep it coarser for computation time

```


Wait, we said KDE was non-parametric, and yet we just defined parameters. 
What does it mean to be "non-parametric"? Does it mean NO parameters?

Non-parametric means no assumed distribution about where the points lie

We need to setup a grid first. We'll use the California shape to get started
```{r}
grid_ca <- ca %>%
  create_grid_rectangular(cell_size = cell_size, side_offset = band_width)
```


Plot the grid just to check
```{r}
tm_shape(grid_ca) + tm_polygons()
```

Let's do the KDE, might take a while to run

```{r}
kde <- ks %>%
  SpatialKDE::kde(band_width = band_width, kernel = "quartic", grid = grid_ca)
```

Visualize the KDE

```{r}
tm_shape(kde) +
  tm_polygons(col = "kde_value", palette = "viridis", title = "KDE Estimate") 
```
A bit hard to see, so how would we allow for interactive mapping?
 
```{r}
tmap_mode("view")

tm_shape(kde) +
  tm_polygons(col = "kde_value", palette = "viridis", title = "KDE Estimate") 
```



### An alternative KDE method using a raster data model

In this first example, we the grid function created a bunch of small square polygons. They LOOK like a raster image, but they're not. 

So for the sake of illustration, let's instead do a KDE with a raster.

We first prepare a raster with given cell size. The raster should be slightly bigger than convex hull of the data.
```{r}
raster_ks <- ks %>%
  create_raster(cell_size = cell_size, side_offset = band_width)
```

Create the KDE (again, this may take some time)
```{r}
kde <- ks %>%
  kde(band_width = band_width, kernel = "triweight", grid = raster_ks)
```


And like before, we can plot it
```{r}
tmap_mode("view")

tm_shape(kde) +
  tm_raster(palette = "viridis", title = "KDE Estimate") +
  tm_shape(ks) 
```

## IDW

Inverse Distance Weighting (IDW) is an interpolation technique that estimates values at unsampled locations based on the weighted average of nearby known values, with closer points having greater influence. 

Let's first start by making another grid
```{r}
grd_idw <- ca %>% 
  st_bbox(.) %>%
  st_as_stars(dx = 2500) %>%
  st_crop(., ca)


grd_idw
```
Here, we chose grid cells not too fine, so that we still see them in plots.

Perhaps the simplest interpolation method is inverse distance weighted interpolation, which is a weighted average, using weights inverse proportional to distances from the interpolation location.


Before we start, let's break down the function and notation
```{r}
idw_result <- gstat::idw(ks$pbe1415~1, ks, grd_idw)
```



In R, the ~1 notation is commonly used in modeling functions to specify an intercept-only model, meaning that no predictor variables are included. This notation appears in various contexts, including spatial interpolation `(gstat::idw())`, `regression (lm())`, and mixed-effects modeling `(lme4::lmer())`, where it indicates that predictions or estimates should be based only on a constant value (e.g., a global mean or spatial trend) rather than explanatory variables.

Let's plot with ggplot instead of tmap

```{r}
ggplot() + geom_stars(data = idw_result, 
                      aes(fill = var1.pred, x = x, y = y)) + 
    xlab(NULL) + ylab(NULL) +
    geom_sf(data = st_cast(ca, "MULTILINESTRING")) #+
    #geom_sf(data = ks)
```
## Questions for thought

1. How does Kernel Density Estimation (KDE) differ from Inverse Distance Weighting (IDW) in terms of what they measure and how they are used?  

2. Why does IDW assume that nearby points have more influence than distant ones, and how might this assumption be problematic in some cases?  

3. KDE produces a smooth density surface, while IDW creates an interpolated surface—how does this difference affect their applications in spatial analysis?  

4. What are some limitations of IDW when interpolating spatial data, especially in areas with sparse or unevenly distributed points?  

## TRY WITH A PARTNER (in-class)

- How does the IDW output change if you change the bandwidth/grid size? Why does it change in the way it does?
- Try mapping a different variable. How would you do this?


```{r}

#New grid

grd_idw <- ca %>% 
  st_bbox(.) %>%
  st_as_stars(dx = 2500) %>%
  st_crop(., ca)

grd_idw

idw_result <- gstat::idw(ks$pctWht0102~1, ks, grd_idw)

ggplot() + geom_stars(data = idw_result, 
                      aes(fill = var1.pred, x = x, y = y)) + 
    xlab(NULL) + ylab(NULL) +
    geom_sf(data = st_cast(ca, "MULTILINESTRING")) #+
    #geom_sf(data = ks)


```





