---
title: "inclass_geometry"
author: "Patrick Bitterman"
date: "2025-03-06"
output: html_document
---


```{r}
library(tidyverse)
library(sf)
library(tmap)
```

# read data

QUESTION: WHAT DOES THE 'skip' parameter do, and why is it necessary?

The skip parameter does not load in the first 4 rows as they are not valid data. 

```{r}

counties <- read_sf("./Data/ohio/oh_counties.gpkg")
rivers <- read_sf("./Data/ohio/ohio_rivers.gpkg")


# 7-day snowfall totals for a given day
snow <- read_csv("./Data/snow/oh_snowfall_data_2025-02-28.csv", skip = 4) 



```
Let's turn the tabular data into points

QUESTION: WHAT IS `crs = 4326`? How would you know what `4326` refers to?

The crs is the coordinate system of the data. 4326 refers to WGS84.

```{r}
snowpoints <- st_as_sf(snow, coords = c("Longitude", "Latitude"), crs = 4326)



tm_shape(counties) + tm_polygons() +
  tm_shape(snowpoints) + tm_bubbles()


```

Let's create geometry using centroids
```{r}
centroids <- st_centroid(counties)

tm_shape(counties) + tm_polygons() +
  tm_shape(centroids) + tm_dots()

```

QUESTION: HOW ARE THESE DISTANCE MEASUREMENTS DIFFERENT?


```{r}



portage <- counties %>% dplyr::filter(str_detect(NAME, "Portage"))

cuy <- rivers %>% 
  dplyr::filter(str_detect(assessmentunitname, "Cuya")) %>%
  st_transform(., st_crs(portage))

sf::st_distance(cuy, portage)

sf::st_distance(cuy, portage, which = "Great Circle")
sf::st_distance(cuy, portage, which = "Euclidean")
sf::st_distance(cuy, portage, which = "Hausdorff")
sf::st_distance(cuy, portage, which = "Frechet")


```

Instead of the polygon for Portage County, let's test wtih the centroid of the county

QUESTION(S): Does using the centroid differ from the polygon measurements above? If so, how? Why? Do the types of distance measurements (the `which` parameter) ALSO differ from geometry to geometry? What does this mean you need to consider when measuring distance?
```{r}
portage_centroid <- st_centroid(portage)

sf::st_distance(cuy, portage_centroid)

sf::st_distance(cuy, portage_centroid, which = "Great Circle")
sf::st_distance(cuy, portage_centroid, which = "Euclidean")
sf::st_distance(cuy, portage_centroid, which = "Hausdorff")
sf::st_distance(cuy, portage_centroid, which = "Frechet")

```



# Tasks

1. Reproject the rivers and counties shapes to UTM 17N (You'll need to figure out the EPSG code). Do the distance measurements change? In what way?

2. In the tmap of the snow totals, scale (and/or shade) the tm_bubbles by the snow totals on January 17th. You'll have to deal with an issue with the data. What's the issue? How did you solve it


